Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-19 paper/main.tex
maciej's edit

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
\mengyan{Maciej:how about ours? same approach or better?}

The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
\mengyan{Maciej: can you replace citations like "Reeve \emph{et al.}" to \textcite{Reeve2014}?}
For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g. is that because the data? So I recommend either remove this or explain it somewhere.}
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} . This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .

Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}. 
In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
In each round, we measure the TIR of benchmark RBS as internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, due to being trained on literature data.
As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
Additionally, the data set was relatively small and covered very small part of the possible experimental space.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
\mengyan{Maciej: can you rephrase the last sentence?}
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
\mengyan{explain the exploration and exploitation balance?}

@@

Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
For RBS with high TIR, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
For RBS with medium TIR, a distance of 1 is enough to produce a meaningful increase in TIR.
That means that as the TIR of examined RBSs increases, 
exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
% the algorithm should be ``encouraged" to explore sequences which are more dissimilar to the current candidates.
The low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012},
which indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
In such cases, a directed process should be strongly encouraged.
This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed.
\mengyan{Maciej: can you add one or more citations here?}

@@

\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this). If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}

Taken together our dataset can be seen as a diverse library of well characterised RBS parts.
Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
\mengyan{Maciej: can you explain what is "given samples standard deviation"?}
The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit into our regression method.
We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.

There is a number of issues that we would like to address in the future.
The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
Essentially, there are two possible reasons for this, one due to nature and one is due to the algorithm.
One the one hand, it might be possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
In that case, improves in the algorithm will not help.
On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
% In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
% Another problem we would like to tackle is the not prediction of the TIR of a given RBS.
% In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
There are open questions need to be addressed for applying machine learning in synthetic biology in general.
For example, 
how can we understand the language of RBS sequences better?
Given a small amount of RBS sequences, how can machine learning algorithm provides more accurate predictions and uncertainty measurement? 
How should we choose the parameters which controls the exploration-exploitation rate?

>>>>

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
\mengyan{Maciej:how about ours? same approach or better?}
\maciej{We shouldn't be making comments on ours in the introduction.}
The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} \maciej{Definitely Learn, Test is only about gathering the data and it has its own limitaions}- there is very limited access to methods and software that can improve and understand designs based on the experimental results.
For example, according to Reeve \emph{et al.} \textcite{Reeve2014}there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g. is that because the data? So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} . This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .

Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}. 
In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
In each round, we measure the TIR of benchmark RBS as internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, due to being trained on literature data.
As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
Additionally, the data set was relatively small and covered very small part of the possible experimental space.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
\mengyan{Maciej: can you rephrase the last sentence?}
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the cluster positioned in the top centre labelled as "G rich" has been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
\mengyan{explain the exploration and exploitation balance?}

@@

Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
For RBS with high TIR, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
For RBS with medium TIR, a distance of 1 is enough to produce a meaningful increase in TIR.
That means that as the TIR of examined RBSs increases, 
exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
% the algorithm should be ``encouraged" to explore sequences which are more dissimilar to the current candidates.
The low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012},
which indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
In such cases, a directed process should be strongly encouraged.
This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.

@@

\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this). If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
\maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}

Taken together our dataset can be seen as a diverse library of well characterised RBS parts.
Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given RBS standard deviation (coefficient of variation).
We have decided to sample biological over technical replicates as this gives better representation of how different populations of a given RBS variant should perform.
We also decided to do 6 replicates of each sample rather than the more established 3 because we felt it gives us better representation both in terms of the calculated mean and standard deviation.
The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit into our regression method.
We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.\\

There is a number of issues that we would like to address in the future.
The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
Essentially, there are two possible reasons for this, one due to nature and one is due to the algorithm.
One the one hand, it might be possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
In that case, improves in the algorithm will not help.
On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
Our predictions are robust even though R2 lower than others.
% In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
% Another problem we would like to tackle is the not prediction of the TIR of a given RBS.
% In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
There are open questions need to be addressed for applying machine learning in synthetic biology in general.
For example, 
how can we understand the language of RBS sequences better?
Given a small amount of RBS sequences, how can machine learning algorithm provides more accurate predictions and uncertainty measurement? 
How should we choose the parameters which controls the exploration-exploitation rate?
